Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-25 icfp.tex
big pass, up to 6

We have implemented the Linear Core type system as a plugin for the Glasgow
Haskell Compiler.
GHC plugins allow developers to inspect and modify programs being compiled by
GHC, at different stages of compilation~\cite{10.1145/3331545.3342599}.
%
In particular, for any given Haskell module, Core plugins run for (and receive
as input) every intermediate program produced in the compilation process, i.e.
from desugaring and after each optimising transformation.
%
This implementation further substantiates our claim that Linear Core is
suitable for the intermediate language of an optimising compiler.
%\todo[inline]{Tem de ser anonimizado, provavelmente alterando a bib
  %entry para omitir autores e o proprio link}
The GHC Linear Core plugin~\cite{cite:linear-core-plugin} implements a
typechecker for Linear Core: given a Core program, our plugin typechecks the
linearity of all expressions bound in that program, failing if a linear
resource is not used \emph{exactly once} according to semantic linearity of the
$(\lambda^\pi_\Delta)$ system.
%
% Furthermore, the implementation accepts all example programs from Chapter~\ref{sec:linearity-semantically}
% deemed well-typed by Linear Core, which are marked with a
% \colorbox{notyet}{\notyetcolorname} background.

@@

The implementation of Linear Core as a typechecker does not follow
directly from the description of the type system because Linear Core
is not \emph{syntax-directed}. Specifically, the most challenging
features are splitting linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
Instead of non-deterministically splitting linear resources amongst
sub-derivations, we thread input/output linear resources through each
derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
%
Pattern variables bound in a case alternative, for a scrutinee not in WHNF, are
introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
variable. To use these pattern variables, the scrutinee resources must be first
$Split$ into fragments. Instead of guessing which resources need to be split,
the implementation consumes tagged fragments of a resource \emph{as needed},
i.e., only when consuming a resource whose usage environment has a resource
tagged $K_j$ do we $Split$ that resource on $K$ and consume the $K_j$ fragment.
%
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. %Further research regarding
%this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}

@@

The results of running the Linear Core GHC plugin on large 
libraries focused around linear types are given by
Figure~\ref{fig:core-plugin-res}.
%
%
We compiled the Haskell standard library for programming with linear types,
\texttt{linear-base} (4000 lines, comprising over 100 modules), the
\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines) library, and
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
session-types library, using our plugin.
%
We count the number of programs accepted by our implementation, where
each top-level binding in a module counts as a program, and every such
binding is typechecked once per optimisation pass (i.e., we typecheck
all intermediate programs produced by GHC). The total amount of programs
rejected by the typechecker are counted dually, also across every optimisation
pass, without halting compilation when a program is rejected. We note that our
implementation is not performance conscious, and types every intermediate
program from scratch, instead of preserving information throughout the
pipeline.

@@

{\small
  \input{prototype/core-plugin-results-2}
  }
%
The results indicate our mostly direct implementation of Linear Core is
successful in accepting the vast majority of the thousands of intermediate
programs produced by GHC when compiling libraries that make extensive use of
linear types.

However, Linear Core does not accept every single program produced by the
optimisation pipeline of GHC. Our manual analysis of rejected programs revealed
excellent results. Programs considered ill-typed by the Linear Core
implementation include:
%
\begin{itemize}
    \item Programs we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
        Namely, programs which scrutinize a variable, but then use the variable in the
        case alternatives of the case scrutinizing it. As discussed in
        Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
        understood as linear as long as applications of linear functions binding these
        variables are reduced \emph{call-by-need}, which is not the case in our system.

>>>>

We have implemented the Linear Core type system as a plugin for the Glasgow
Haskell Compiler.
GHC plugins allow developers to inspect and modify programs being compiled by
GHC, at different stages of compilation~\cite{10.1145/3331545.3342599}.
%
In particular, for any given Haskell module, Core plugins run for (and receive
as input) every intermediate program produced in the compilation process, i.e.
from desugaring and after each optimising transformation.
%
Our implementation further substantiates our claim that Linear Core is
suitable for the intermediate language of an optimising compiler.
%\todo[inline]{Tem de ser anonimizado, provavelmente alterando a bib
  %entry para omitir autores e o proprio link}
The GHC Linear Core plugin~\cite{cite:linear-core-plugin} implements a
typechecker for Linear Core: given a Core program, our plugin typechecks the
linearity of all expressions bound in that program, failing if a linear
resource is not used \emph{exactly once} according to semantic linearity of the
$(\lambda^\pi_\Delta)$ system.
%
% Furthermore, the implementation accepts all example programs from Chapter~\ref{sec:linearity-semantically}
% deemed well-typed by Linear Core, which are marked with a
% \colorbox{notyet}{\notyetcolorname} background.

@@

The implementation of Linear Core as a typechecker does not follow
directly from the description of the type system because Linear Core
is not \emph{syntax-directed}. Specifically, the most challenging
features are splitting linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
Instead of non-deterministically splitting linear resources amongst
sub-derivations, we thread input/output linear resources through each
derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
%
For a scrutinee not in WHNF, pattern variables in a case alternative are
introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
variable. To use these pattern variables, the scrutinee resources must be first
$Split$ into fragments. Instead of guessing which resources need to be split,
the implementation consumes tagged fragments of a resource \emph{as needed},
i.e., only when consuming a resource whose usage environment has a resource
tagged $K_j$ do we $Split$ that resource on $K$ and consume the $K_j$ fragment.
%
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. %Further research regarding
%this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}

@@

The results of running the Linear Core GHC plugin on large 
libraries focused around linear types are reported in
Figure~\ref{fig:core-plugin-res}.
%
%
We compiled the Haskell standard library for programming with linear types,
\texttt{linear-base} (4000 lines, comprising over 100 modules), the
\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines) library, and
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
session-types library, using our plugin.
%
We count the number of programs accepted by our implementation, where
each top-level binding in a module counts as a program, and every such
binding is typechecked once per optimisation pass (i.e., we typecheck
all intermediate programs produced by GHC). The total amount of programs
rejected by the typechecker are counted dually, also across every optimisation
pass, without halting compilation when a program is rejected. We note that our
implementation is not performance conscious, and types every intermediate
program from scratch, instead of preserving information throughout the
pipeline.

@@

{\small
  \input{prototype/core-plugin-results-2}
  }
%
The results indicate that our mostly direct implementation of Linear Core is
successful in accepting the vast majority of the thousands of intermediate
programs produced by GHC when compiling libraries that make extensive use of
linear types.
%
However, Linear Core does not accept every single program produced by the
optimisation pipeline of GHC. Our manual analysis of rejected programs revealed
excellent results. Programs considered ill-typed by the Linear Core
implementation include:
%
\begin{itemize}
    \item Programs we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
        Namely, programs which scrutinize a variable, but then use the variable in the
        case alternatives of the case scrutinizing it. As discussed in
        Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
        understood as linear as long as applications of linear functions binding these
        variables are reduced \emph{call-by-need}, which is not the case in our system.
